---
title: "02 - Exercises: Types of Variables, Loops, and Vectorization"
author: "Diego Mañanes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
documentclass: article
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
classoption: a4paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  dpi = 300,
  fig.align = "center", 
  fig.width = 7, 
  fig.height = 4
)
```

## Variables and basic structures
  
### Assigning Names to Objects

1. Create three variables `x`, `y`, and `z`, with values `10`, `20`, and `30` respectively. Compute their sum and store it in a new variable `total`.
2. Print the class of each variable using `class()`.

```{r}

```


### Working with Vectors

1. Create a vector `v` containing numbers from `5` to `15`.
2. Use vectorized operations to:
  * Add `10` to each element.
  * Calculate the square of each element.
  * Determine which elements are greater than `12`.
3. Create the vector `w <- 1:5` and guess the output of `v + w`


```{r}

```


### Indexing and Subsetting

1. Create a vector `cities` with the elements: `"New York"`, `"London"`, `"Paris"`, `"Berlin"`, `"Palencia"`.
2. Use indexing to:
  * Access the first and third elements.
  * Exclude the second element.
  * Retrieve elements that start with the letter `"P"`. Hint: Use `grep()`

```{r}

```

  
### Logical Operations

1. Create a vector `scores` with values `55, 78, 92, 45, 89`.
2. Use logical operations to:
  * Identify elements greater than `70`.
  * Count how many elements are greater than `70`.
  * Find the position of scores greater than `70`.
  
```{r}

```


### Loops and Vectorization

1. Write a `for` loop to calculate the square of each number in the vector `1:10`. Store the results in a new vector.
2. Rewrite the same operation using vectorized functions.

```{r}

```


### Matrices

1. Create a `3x3` matrix with numbers from `1` to `9`.
2. Use indexing to:
  * Access the element in the second row and third column.
  * Retrieve all elements in the first row.
  * Replace the second column with the vector `10, 20, 30`.
3. Compute the row sums and column sums of the matrix.

```{r}

```


### Functions

1. Write a function `square_and_add()` that takes a numeric vector and a number as input, squares the vector (`^`), adds the number to each element, and returns the result.

```{r, eval = FALSE}
## example
square_and_add(c(1, 2, 3), 10)  # Expected: 11, 14, 19
```

2. Test your function with different inputs.

```{r}

```


### Factors

1. Create a factor `colors` from the vector `c("red", "blue", "red", "green", "blue")` with levels `"red"`, `"blue"`, and `"green"`.
2. Check the class of the factor and its levels.

```{r}

```


### Data Frames 

1. Create a data frame `df` with three columns:
  * `Name`: `"Alice", "Bob", "Charlie"`
  * `Age`: `25, 30, 35`
  * `Score`: `80, 95, 90`
2. Access:
  * The second row.
  * The `Score` column.
  * The `Age` of `"Bob"`.

```{r}

```


### Working with Lists

Create a list that contains the following elements:

1. A numeric vector with the numbers 1 to 10.
2. A character vector with the names of the first three months of the year.
3. A matrix with 6 rows and 2 columns filled with random numbers between 1 and 100.
4. Another list that contains:
   * A factor with the levels "Yes" and "No".
   * A logical vector of length 5.

```{r}

```

   
Now, do the following tasks: 

1. Extract the numeric vector from the list and compute its sum.
2. Retrieve the second column of the matrix in the list and find the maximum value.
3. Add a new element to the main list: a data frame with columns `ID` (1 to 5) and `Score` (random numbers between 50 and 100).
4. Access the factor and check its levels.

```{r}

```


### Manipulating Data Frames

Create a data frame with the following columns:
1. `Name`: A character vector with 6 names.
2. `Age`: A numeric vector with random ages between 20 and 40.
3. `Score`: A numeric vector with random scores between 50 and 100.
4. `Passed`: A logical vector indicating if the score is greater than or equal to 50.

```{r}

```


Now, do the following:

1. Subset the data frame to include only rows where `Passed` is `TRUE`.
2. Add a new column called `Category`:
   * Assign "Young" if `Age` < 30.
   * Assign "Middle-aged" otherwise.
3. Compute the average score of people in the "Young" category.
4. Sort the data frame by `Score` in descending order.

```{r}

```


## Programming basics

### Conditional Expressions

1. Write a function that receives a numeric value between -10 and 40 and by using a conditional statement:
   * Prints "Below freezing" if the temperature is below 0°C.
   * Prints "Cold" if the temperature is between 0°C and 10°C.
   * Prints "Warm" if the temperature is between 10°C and 25°C.
   * Prints "Hot" if the temperature is above 25°C.

```{r}

```

### Creating Functions

1. Create a function called `grade_classifier` that:
   - Accepts a numeric score between 0 and 100.
   - Returns:
     - "Fail" for scores below 50.
     - "Pass" for scores between 50 and 74.
     - "Distinction" for scores of 75 or more.

```{r}

```


### Scoping

1. Write a function `outer_function` that:
   * Defines a variable `x` inside the function as 5.
   * Contains a nested function `inner_function` that defines a variable `y` as 10.
   * Returns the sum of `x` and `y` using `inner_function`.

  Call `outer_function` and check whether `x` and `y` are accessible in the global environment.

2. Create a function `scope_test` that:
   * Takes no arguments.
   * Modifies a global variable `counter` by increasing it by 1.
   * Returns the updated value of `counter`.

   Test this function with the following steps:
    * Define `counter <- 0` in the global environment.
    * Call `scope_test` multiple times and observe how the value of `counter` changes.

```{r}

```

## For loops and `apply`-related functions

### Loops vs Vectorization

1. Write a `for` loop that calculates the square of numbers from 1 to 10 and stores the results in a vector called `squares`. Rewrite the same code using vectorized operations.  

2. Rewrite the following `while` loop using vectorization:  

```{r}
i <- 1
results <- c()
while (i <= 10) {
  results <- c(results, i^2)
  i <- i + 1
}
results
```


### `apply`

1. Given the following matrix:  

```{r}
(mat <- matrix(1:12, ncol = 3))
```

* Use `apply` to compute the sum of each row.  
* Use `apply` to compute the product of each column.  

2. Create a 5x5 matrix of random numbers between 1 and 50 (check `sample`). Use `apply` to calculate the mean of each row and adding 5 to these values.



### Aggregation with `aggregate()` and `by()`

Using the following data frame:  

```{r}
(df <- data.frame(
  age = c(21, 25, 19, 23, 22, 29, 20), 
  gender = factor(c("male", "female", "female", "male", "female", "male", "male"))
))
```

1. Use `aggregate()` to compute the mean and standard deviation of `age` grouped by `gender`.  

```{r}

```


2. Use the `by()` function to calculate the mean and range of `age` grouped by `gender`.

```{r}

```


3. Add a new column `score` to the data frame. Use `aggregate()` to compute the mean of both `age` and `score` grouped by `gender`.  

```{r}

```



### Using `split()` and `lapply()`  

Split the following data frame by the `group` column:  

```{r}
(data <- data.frame(
  value = c(5, 10, 15, 20, 25, 30, 35),
  group = factor(c("A", "A", "B", "B", "C", "C", "C"))
))
```

Use `lapply` to calculate the sum of `value` for each group:

```{r}

```


### Function for `lapply`

Given this list: 

```{r}
data_list <- list(
  df1 = data.frame(A = c(1, 2, 3), B = c(4, 5, 6)),
  df2 = data.frame(A = c(10, 20, 30), C = c(40, 50, 60)),
  df3 = data.frame(B = c(100, 200, 300), A = c(1, 2, 3))
)
data_list
```

Create a function that receives two parameters: a list and a string. Inside the function, it should iterate the list (using `lapply`) and compute the mean of the column given in the second parameter.

```{r}

```

