---
title: "03 - Exercises: Data visualization"
author: "Diego Mañanes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
documentclass: article
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
classoption: a4paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  dpi = 300,
  fig.align = "center", 
  fig.width = 7, 
  fig.height = 4
)
```

## Using R base

You will have to create different plots using the following mock data: 

```{r}
set.seed(123)
df.data <- data.frame(
  X = rnorm(1000, mean = 1000, sd = 10),
  Y = rpois(100, n = 1000),
  Category.1 = factor(sample(LETTERS[1:3], size = 1000, replace = T)),
  Z = rnorm(1000, mean = 12, sd = 6)
)
```


### Scatter Plots

1. Create a scatter plot using the `df.data` dataset:
   * Plot `X` on the x-axis and `Z` on the y-axis.
   * Use different colors for points based on the `Category.1` variable.
   * Add a legend to indicate the categories.

```{r}

```


2. Add horizontal and vertical lines to the plot:
   * Add a vertical line at `X = 1000`.
   * Add a horizontal line at `Z = 12`.

```{r}

```


### Histograms

1. Create a histogram of the `Y` variable from the `df.data` dataset:
   * Use 30 bins and set the color to "purple".
   * Add appropriate labels for the x-axis, y-axis, and a title.

```{r}

```


2. Create a histogram for the `X` variable:
   * Use 20 bins and set the color to "orange".
   * Overlay a density line on the histogram.

```{r}

```


### Boxplots and barplots

1. Create a boxplot for `Y` grouped by `Category.1`:
   * Use different colors for each category.
   * Add a title and axis labels.

```{r}

```


2. Create a barplot for the first 15 samples of `X`:
   * Use `paste0("Sample-", 1:15)` as the names for the bars.
   * Use colors based on the `Category.1` variable.

```{r}

```


### Combined Plots

1. Use the `par()` function to display:
   * A histogram of `Y` in the left panel.
   * A boxplot of `Z` grouped by `Category.1` in the right panel.
   * Ensure each plot has a title and axis labels.

```{r}

```


2. Create a 2x2 grid of plots:
   * Top-left: Histogram of `X`.
   * Top-right: Scatter plot of `X` vs. `Z`.
   * Bottom-left: Boxplot of `Y` grouped by `Category.1`.
   * Bottom-right: Barplot for the first 10 samples of `X`.

```{r}

```


## Using `ggplot2`

Here are several exercises based on the `ggplot2` contents. These exercises are progressive, ranging from basic to advanced levels, and aim to reinforce understanding of the concepts.

```{r}
data(iris)
```


### Basic plots

1. Basic scatter plot. Using the `iris` dataset. Create a scatter plot of `Sepal.Length` vs. `Sepal.Width`:  

```{r}

```


2. Color mapping. Modify the previous scatter plot:  
   * Map the color of the points to the `Species` column.  
   * Add a title and label the axes.  

```{r}

```


3. Basic histogram
   * Create a histogram of the `Petal.Length` column.  
   * Set the number of bins to 30.

```{r}

```



### Adding elements

1. Faceting:
    * Create a scatter plot of `Sepal.Length` vs. `Sepal.Width` with points colored by `Species`.  
    * Create separate panels for each species.  

```{r}

```


2. Adding a regression line:
   * Create a scatter plot of `Sepal.Length` vs. `Sepal.Width`.  
   * Add a linear regression line using `geom_smooth(method = "lm")`.  
   * Group the regression lines by `Species`.  

```{r}

```


3. Boxplot with custom colors:
   * Create a boxplot of `Petal.Width` grouped by `Species`.  
   * Use custom colors (`gold`, `lightblue`, `orange`) for the boxes.  
   * Add a horizontal line at `y = 1.5` with a dashed red line.  

```{r}

```


4. Density plot:
   * Create a density plot of `Petal.Length`.  
   * Fill the plot by `Species` and set `alpha = 0.5`.  
   * Change the theme to `theme_minimal()`.  

```{r}

```


### More elements

1. Complex plot with multiple layers:
   * Create a scatter plot of `Sepal.Length` vs. `Sepal.Width` with points colored by `Species`.  
   * Add a smooth line (`geom_smooth()`) for the overall trend (not grouped).  
   * Use `facet_wrap()` to create separate panels for each species.  
   
```{r}

```


2. Using continuous variables:
   * Create a scatter plot of `Petal.Length` vs. `Petal.Width`.  
   * Map the color of the points to `Sepal.Length` and the size of the points to `Sepal.Width`.  
   * Add axis labels and a title to the plot.  

```{r}

```


3. Custom theme and labels:
    * Create a boxplot of `Petal.Length` grouped by `Species`.  
    * Change the plot's theme to `theme_classic()`.  
    * Customize the axis labels and title with `labs()` and modify the title text style using `theme()`.  

```{r}

```


4. Combining geometries:
    * Create a histogram of `Petal.Length` filled by `Species`.  
    * Overlay a density plot for the same variable with `geom_density()`.  
    * Use custom colors for the density plot fills.  

```{r}

```

