---
title: "04 - Exercises: Brief introduction to tidyverse"
author: "Diego Mañanes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
documentclass: article
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
classoption: a4paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  dpi = 300,
  fig.align = "center", 
  fig.width = 7, 
  fig.height = 4
)
```


## 1. Tidy data

Explain why the following data is not tidy and restructure it into a tidy format.

**Hint:** Use the `tidyr` package and functions like `pivot_longer()` or the `melt()` function from the `reshape2` pacakge. 

```{r}
data <- data.frame(
  Name = c("Alice", "Bob"),
  Subject1_Score = c(85, 88),
  Subject2_Score = c(90, 72),
  Subject3_Score = c(78, 95)
)
data
```
```{r}

```


## 2. Using Pipes

Rewrite the following code using pipes:

```{r}
  result <- head(iris)
  result <- result[order(result$Sepal.Length, decreasing = TRUE), ]
  result <- result[result$Species == "setosa", ]
```

```{r}

```


## `filter()` function

Filter the `iris` dataset to:

* Include only rows where `Sepal.Width > 3`.
* Exclude rows where the species is "virginica".
* How many rows remain after applying these filters?

```{r}
data(iris)
```


## `arrange()` 

Sort the `iris` dataset:

* By `Petal.Length` in descending order.
* If two rows have the same `Petal.Length`, sort them by `Sepal.Width` in ascending order.

```{r}

```


## `mutate()`

* Add a new column `Sepal.Area` (product of `Sepal.Length` and `Sepal.Width`).
* Categorize `Sepal.Area` into "Large" (>= 20) and "Small" (< 20).

```{r}

```


## `group_by()` and `summarize()`

Group the `iris` dataset by `Species` and calculate:

* The average `Sepal.Length` and `Petal.Length`.
* The total number of samples per species.

```{r}

```


## Joining Data Frames

Consider the following datasets:

`df1`:

| ID   | Value |
|------|-------|
| ID_1 | 10    |
| ID_2 | 20    |
| ID_3 | 30    |

`df2`:

| ID   | Category |
|------|----------|
| ID_1 | A        |
| ID_2 | B        |
| ID_4 | C        |

* Perform the following joins using `left_join()`, `right_join()`, and `inner_join()`. 
* Explain the differences in the results.

```{r}

```


## Combination of tasks

1. Filter the `iris` dataset to include only rows where `Petal.Length > 1.5`.
2. Add a new column `Sepal.Ratio` (ratio of `Sepal.Length` to `Sepal.Width`).
3. Group the data by `Species` and calculate the average `Sepal.Ratio`.
4. Arrange the results in descending order of the calculated average.

```{r}

```


## Analyzing the `nycflights13` data

Use the `nycflights13::flights` dataset to answer the following:

1. Which airline (`carrier`) had the most flights in 2013? Combine the `flights` dataset with the `airlines` dataset to replace `carrier` codes with their full names.
2. For each destination (`dest`), calculate the average arrival delay.

```{r}

```


