---
title: "More exercises"
author: "Diego Mañanes"
date: '`r strftime(Sys.time(), format = "%B %d, %Y")`'
documentclass: article
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
classoption: a4paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  dpi = 300,
  fig.align = "center", 
  fig.width = 7, 
  fig.height = 4
)
projectPath <- here::here()
setwd(file.path(projectPath, "01-R-basics/data")) ## what is this???

library("dplyr")
```


## Introduction 

These exercises don't belong to a particular module, the idea is to solve them using the tools we have learned. 

## Leukemia dataset

### Loading data

You need to read the `leukemia.data.txt` and `leukemia.class.txt` files located in the `data` folder. 

* `leukemia.data.txt`: make sure you read it correctly: the file contains row names and we don't want them to be part of the data frame content. Hint: check the `row.names` parameter from the `read.table` function. In addition, this file contains just numbers, so in order to be able to apply functions such as `apply()`, it is better if you transform it to a matrix (check `as.matrix()` or `data.matrix()` functions).
* `leukemia.class.txt`: we could read it as a data frame. However, as it is just a vector, it is easier if we use the `scan()` function. Check the documentation and the `what` parameter. This should be encoded as a factor. 
* Create a vector with the sex of patients. Do it by alternating the labels `"Male"` and `"Female"` and convert the object into a factor. * Finally, bind the two factors into a data frame. 

```{r}

```

### Counting samples

Now, just count the number of patients that we have for each categorical variable using the `table()` function.

```{r}

```

Do the same using the `group_by()` and `summarize()` functions + the `n()` function. 

```{r}

```


### Accessing data

Let's compute different stuff for different patients. Assume that `leuk.dat.m` has patients in columns and genes in rows. 

* Compute the mean expression levels of all genes together only for those patients that are Female.
* Same as above but for every gene separately. 
* Compute the median expression levels of every gene grouped by Sex. 
* Use the `t.test()` function to compute those genes that are different between the class of leukemia. then, obtain the gene name of those with p-value <= 0.001. 
* Compute the mean of the second gene for males and females. 

```{r}

```


```{r}

```


```{r}

```


```{r}

```



